\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-11-04 paper.tex
update simulated data generation process

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top\]
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

Now, concerning the simulation of the longitudinal data, the idea is to sample from multivariate normal distributions. 
We use linear time-varying features such that
\[ Y_i^l(t) = \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \big( (1, t)^\top \beta_k^l + (1, t)^\top b_i^l + \epsilon_i^l(t) \big) \]
where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to 
\[ \beta_k^l \sim \ind{\{l \in \cS_k \}} \cN\Big( 
\mu_k, 
\begin{bmatrix}
  \rho_3 & 0\\
  0 & \rho_3
\end{bmatrix} 
\Big)
+ \ind{\{l \notin \cS_k \}} \cN\Big( 
\begin{pmatrix}
  0 \\ 0
\end{pmatrix}
, 
\begin{bmatrix}
  \rho_4 & 0\\
  0 & \rho_4
\end{bmatrix} 
\Big) \]
for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that $\beta_k^l$ coefficients are both positive for high-risk subjects ($G_i=1$), which favours increasing trajectories with positive values, and negative otherwise ($G_i=0$), which favours decreasing trajectories with negative values. This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.

@@

To generate survival times, we choose a risk model of the form
\begin{align*}
\lambda_i(t|G_i = k) = \lambda_0(t) \exp \Big\{ \sum_{l=1}^L &\gamma_{k,1}^l (\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t) \\
&+ (\gamma_{k,2,1}^l b_{i,1}^l + \gamma_{k,2,2}^l b_{i,2}^l) + \gamma_{k,3}^l (\beta_{k,2}^l + b_{i,2}^l) \Big\},
\end{align*}
which corresponds to~\eqref{eq:intensity-model} taking the first three functionals described in Table~\ref{table:shared_associations} as shared associations.
Now, for the choice of the association parameters, we want to induce sparsity (similarly to~\eqref{eq:xi-sparse}), so we set 
\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] and
\begin{equation}
  \label{eq:gamma-sparse}
  \gamma_{k,a}^l = \varsigma_2 \big(\ind{\{l \leq (k + 1) \lfloor L r_s/K \rfloor \}} + \ind{\{l \in \cS_k \}} \big) \mathbf{1}_{\imath_a},
\end{equation}
so that there is a proportion $r_s$ of active longitudinal features among the $L$, and the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$.
Note that one can write
\[ \lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\},\]
being a Cox model with a linear time-varying feature.
We choose a Gompertz distribution~\citep{gompertz1825xxiv} for the baseline, that is
\begin{equation}
  \label{eq:baseline}
  \lambda_0(t) = \kappa_1 \kappa_2 \exp(\kappa_2t)
\end{equation}
with $\kappa_1 > 0$ and $\kappa_2 \in \R$ the scale and shape parameters respectively, which is a common distribution choice in survival analysis~\citep{klein2005survival} with a rich history in describing mortality curves. One can now generate survival times explicitly as
\begin{equation}
  \label{eq:time-generation}
  T_i^\star | G_i=k \sim \dfrac{1}{\iota_{i,k,2} + \kappa_2} \log \Big(1 - \dfrac{(\iota_{i,k,2} + \kappa_2) \log U_i}{\kappa_1 \kappa_2 \exp\iota_{i,k,1}} \Big)
\end{equation}
where $U_i \sim \cU\big([0,1]\big)$ (see Appendix~\ref{sec:event-time-simu} for the derivation of this expression).
The distribution of the censoring variable $C_i$ is the geometric distribution $\cG(\alpha_c)$, where $\alpha_c \in (0, 1)$ is empirically tuned to maintain a desired censoring rate $r_c \in [0,1]$. 
Finally, the prediction time for subject $i$ is draw according to 
\begin{equation}
  \label{eq:law-t_i^max}
  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
\end{equation}
with $(\nu_1, \nu_2) \in \R_+^2$ two shape parameters. Figure~\ref{fig:timax-pdf} shows the corresponding probability density function for $t_i^{max}$ with some explanations.
We then truncate all $Y_i$ measurements being after $t_i^{max}$, since it represents the ``present'' time, so the time up to which one has longitudinal data.
The choice of all hyper-parameters is driven by the applications on real data presented in Section~\ref{sec:application}, and summarized in Table~\ref{table:parameters choice}. Figure~\ref{fig:example-data} gives an example of data generated according to the design we have just described.
\begin{table}[htb]
\caption{Hyper-parameter choices for simulation. Let us also precise that we take $n \in [200, 4000]$ and $(p, L) \in [3, 300]^2$.
\label{table:parameters choice}}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{ccccccccccccc}
\toprule
$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\nu_1, \nu_2)$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
\midrule
0.4 & $(, 10^{-3},)$ & 
$\begin{pmatrix}
  1 \\ 0.3
\end{pmatrix}$ & 
$\begin{pmatrix}
  1 \\ 0.5
\end{pmatrix}$ 
& 0.5 & 4 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
\bottomrule
\end{tabular}}
\end{table}
\begin{figure}[!htb]
\centering
\subfigure[Density of the law used to simulate $t_i^{max}$ according to~\eqref{eq:law-t_i^max} with $(\nu_1, \nu_2) = (2,5)$, to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.]{
\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
\label{fig:timax-pdf}}
\hspace{.02\textwidth}
\subfigure[exemple simu]{
% \includegraphics[width=.38\textwidth]{./figures/time2.pdf}
\label{fig:example-data}}
\caption{Illustrations of simulated data.}
\end{figure}

>>>>

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\begin{equation}
  \label{eq:alpha-def}
  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
\end{equation}
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

Now, concerning the simulation of the longitudinal data, the idea is to sample from multivariate normal distributions. 
We use linear time-varying features such that
\[ Y_i^l(t) = \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \big( (1, t)^\top \beta_k^l + (1, t)^\top b_i^l + \epsilon_i^l(t) \big) \]
where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to 
\[ \beta_k^l \sim \cN\Big( 
\mu_k, 
\begin{bmatrix}
  \rho_3 & 0\\
  0 & \rho_3
\end{bmatrix} 
\Big) \]
for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that trajectories start on average higher for high-risk subjects ($G_i=1$), and with a higher slope than low-risk subjects ($G_i=0$). This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.

@@

Now to generate survival times, we want to induce sparsity in the longitudinal data effects, so we set 
\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] 
and we choose a risk model~\eqref{eq:intensity-model} with
\begin{equation*}
  \left\{
    \begin{array}{ll}
    \big(\varphi_{k,a}(t, b_i^l)\big)_{a \in \{1,2\}} = \big(\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t\ , {b_i^l}^\top \big)^\top \ind{\{l \in \cS_k\}} + \cN \big(0, \bSigma_3(\rho_3)\big) \ind{\{l \notin \cS_k\}}  \\
    \big( \varphi_{k,a}(t, b_i^l)\big)_{a=3,\ldots,\cA} = \cN \big(0, \bSigma_4(\rho_3)\big),
    \end{array}
  \right.
\end{equation*}
and 
\begin{equation}
  \label{eq:gamma-sparse}
  \gamma_{k,a}^l = \varsigma_2 \mathbf{1}_{\imath_a} \ind{\big\{l \in \cS_k,\ a\in \{1, 2\} \big\}}
\end{equation}
for the choice of the association parameters, with $\cA = 3 + 3(1-r_s)/r_s$ so that $(i)$ there is a proportion $r_s$ of active longitudinal features among the $L$ (with the $\cS_k$), $(ii)$ we induce sparsity within each group adding a proportion of $(1-r_s)$ non-active association features ($a \geq 3$) in each group (similarly to~\eqref{eq:xi-sparse} for the time-independent features), and $(iii)$ the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$ (thanks to the $\cS_k$ definition).

In terms of active association features, it corresponds to~\eqref{eq:intensity-model} taking the first two functionals described in Table~\ref{table:shared_associations} as shared associations with $\alpha = 1$ in~\eqref{eq:alpha-def} so that one can write
\[ \lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\},\]
being a Cox model with a linear time-varying feature that allows the following explicit survival times generation process. Then, we do not consider the time-dependent slope as association feature since here with $\alpha = 1$, it would be collinear with the random effects.
We choose a Gompertz distribution~\citep{gompertz1825xxiv} for the baseline, that is
\begin{equation}
  \label{eq:baseline}
  \lambda_0(t) = \kappa_1 \kappa_2 \exp(\kappa_2t)
\end{equation}
with $\kappa_1 > 0$ and $\kappa_2 \in \R$ the scale and shape parameters respectively, which is a common distribution choice in survival analysis~\citep{klein2005survival} with a rich history in describing mortality curves. One can now generate survival times explicitly as
\begin{equation}
  \label{eq:time-generation}
  T_i^\star | G_i=k \sim \dfrac{1}{\iota_{i,k,2} + \kappa_2} \log \Big(1 - \dfrac{(\iota_{i,k,2} + \kappa_2) \log U_i}{\kappa_1 \kappa_2 \exp\iota_{i,k,1}} \Big)
\end{equation}
where $U_i \sim \cU\big([0,1]\big)$ (see Appendix~\ref{sec:event-time-simu} for the derivation of this expression).
The distribution of the censoring variable $C_i$ is the geometric distribution $\cG(\alpha_c)$, where $\alpha_c \in (0, 1)$ is empirically tuned to maintain a desired censoring rate $r_c \in [0,1]$. 
Finally, the prediction time for subject $i$ is draw according to 
\begin{equation}
  \label{eq:law-t_i^max}
  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
\end{equation}
with $(\nu_1, \nu_2) \in \R_+^2$ two shape parameters. Figure~\ref{fig:timax-pdf} shows the corresponding probability density function for $t_i^{max}$ with some explanations.
We then truncate all $Y_i$ measurements being after $t_i^{max}$, since it represents the ``present'' time, so the time up to which one has longitudinal data.
The choice of all hyper-parameters is driven by the applications on real data presented in Section~\ref{sec:application}, and summarized in Table~\ref{table:parameters choice}. Figure~\ref{fig:example-data} gives an example of data generated according to the design we have just described.
\begin{table}[htb]
\caption{Hyper-parameter choices for simulation. Let us also precise that we take $n \in [200, 4000]$ and $(p, L) \in [3, 300]^2$.
\label{table:parameters choice}}
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{ccccccccccccc}
\toprule
$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\nu_1, \nu_2)$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
\midrule
0.4 & $(, 10^{-3},)$ & 
$\begin{pmatrix}
  1 \\ 0.3
\end{pmatrix}$ & 
$\begin{pmatrix}
  1 \\ 0.5
\end{pmatrix}$ 
& 0.5 & 4 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
\bottomrule
\end{tabular}}
\end{table}
\begin{figure}[!htb]
\centering
\subfigure[Density of the law used to simulate $t_i^{max}$ according to~\eqref{eq:law-t_i^max} with $(\nu_1, \nu_2) = (2,5)$, to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.]{
\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
\label{fig:timax-pdf}}
\hspace{.02\textwidth}
\subfigure[exemple simu]{
% \includegraphics[width=.38\textwidth]{./figures/time2.pdf}
\label{fig:example-data}}
\caption{Illustrations of simulated data.}
\end{figure}
